Performance benefits of Intel\textsuperscript{\textregistered}~Optane\texttrademark~DC persistent memory for the parallel processing of large neuroimaging data

### 2019-12-17 paper.tex
some fixes

\section{Introduction}
Neuroimaging open-data initiatives have led to extensively large repositories of
publicly available data. Such initiatives include the \bigbrain~\cite{BigBrain}, 
a one-of-a-kind $\SI{603}{\gibi\byte}$
histological image of a 65 year-old healthy human brain
at 20$\mu$m resolution; the UK Biobank~\cite{ukbiobank}, a repository expected to
contain approximately $\SI{0.2}{\peta\byte}$ of data (including various magnetic
resonance (MR) imaging modalities) from 500,000 individuals living in the UK;
the Human Connectome Project~\cite{HCP}, a repository containing MR scans from
1,200 healthy adults, expected to exceed $\SI{2}{\peta\byte}$ in size; 
and the Consortium for Reliability and Reproducibility (CoRR)~\cite{corr}, an
initiative which aggregates MR data from various centres around the world, 
of which 32 are currently available and make up about $\SI{937}{\giga\byte}$ of
data in total.

@@

Intel Optane DC Persistent Memory Module~\cite{optanebrief} (DCPMM) is a high-performance
storage technology that resides on the DIMMs to reduce latency to the device.
The first generation of Intel Optane DCPMM offers capabities of 128~GB,
256~GB, and 512~GB, enabling it to be more cost effective than DRAM storage.
There are two configuration modes, Memory mode and App Direct mode,
that enable the storage to either be accessed as an extension of volatile main memory
or as a non-volatile memory storage device\footnote{https://itpeernetwork.intel.com/intel-optane-dc-persistent-memory-operating-modes/}.

@@

The server used consisted of 12X~\SI{64}{\gibi\byte} DRAM devices,
resulting in a total of \SI{768}{\gibi\byte} of DRAM, and 12X~\SI{256}{\giga\byte} Optane DCPMM
modules, resulting in total of \SI{3}{\tebi\byte} of Intel Optane DC persistent memory.
Other storage devices included a \SI{240}{\gibi\byte} Micron SATA SSD, of which
only \SI{149}{\gibi\byte} where available for the experiments, and a
\SI{720}{\tebi\byte} shared Dell EMC Isilon network-attached storage platform. Isilon is an nsf4 mounted
cluster (10Gb network) consisting of 5 nodes of 36 hard disk drives (HDD). The local disk
was the only storage device set up as a writeback device. Both Isilon and Optane DCPMM in
App Direct mode were configured as write-through devices, meaning they did not leverage
DRAM as a cache for writes. For Optane DC persistent memory, this was achieved by configuring the xfs filesystem with
direct access (DAX). DAX enables persistent memory to be byte addressable, bypassing the page cache. All storage devices were benchmarked using the following script: \url{https://github.com/ValHayot/paper-memory-storage/scripts/bench_disks.sh}.
The result of the benchmarks can be seen in Table \ref{table:bandwidths}.

@@

Memory mode leverages Intel Optane DC persistent memory (DCPMM) to extend the system's available
memory. In this mode, Optane DCPMM uses DRAM as a cache and is accessible as
volatile addressable main memory. By extending main memory, Memory mode enables the
fast access of large volumes of data. For instance, Memory mode on our server enabled use of
3~TiB of main memory, whereas it would have only had 768~GiB of main memory if Optane DCPMM could not be
leveraged. Although DRAM is used as cache by Optane DCPMM, it is not visible to the operating system.

@@

App Direct mode enables Optane DCPMM to be accessed as a high-performance storage device.
Unlike Memory mode, the OS is able to differentiate between DRAM and Optane DCPMM,
treating them as two distinct memory tiers. Optane DCPMM
does not use the DRAM as cache in App Direct Mode when configured with DAX, as is our case.

Similarly to Memory Mode, our experiments evaluated all available filesystems in App Direct
Mode. In other words, we evaluated DRAM (tmpfs mount), Optane DCPMM (\SI{1.5}{\tebi\byte} persistent
memory mount), local SSD and Isilon.

@@

For our experiments, we used the entire CoRR dataset, made available on \href{https://www.datalad.org/}{DataLad} and
applied Participant analysis to it. The BIDS App Example was executed using a Singularity container stored on
Isilon. As in the BigBrain Incrementation, the experiment was parallelized 
using GNU Parallel with 25 and 96 processes. The conditions were executed in both App Direct and Memory Mode,
where Isilon and the local SSD were only evaluated in Memory Mode. Each experiment was repeated 3x.

@@

\subsection{40~$\mu$m \bigbrain incrementation}
\subsubsection{25 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-40bb_25cpus.pdf}
    \captionsetup{belowskip=-10pt}
    \caption{Makespan of the \bigbrain Incrementation application using 25 
             processes on the 125 blocks on the 40~$\mu$m image given different
             storage configurations. Anticipated read and write duration was 
             measured by applying the perceived bandwidths Table~\ref{table:bandwidths}
             to Eq.~\ref{eq:makespan}. Three 
             repetitions were performed. *Local SSD did not complete the
         writing of the last few blocks (approximately 5-10) due to storage limitations.}\label{fig:makespan-25cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-40bb_25cpus.pdf}
    \captionsetup{belowskip=0pt}
    \caption{Total read/increment/write breakdown of \bigbrain Incrementation using 25 cpus. Three repetitions were performed. *Local SSD did not
    complete the writing of the last few blocks (approximately 5-10) due to storage limitations}\label{fig:stacked-25cpus}
\end{figure}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572380543-tmpfsAD_1it_25cpus_40bb-1.pdf}
    \caption{DRAM}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572382482-optaneAD_1it_25cpus_40bb-1.pdf}
    \caption{Optane DCPMM}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572382794-localAD_1it_25cpus_40bb-1.pdf}
    \caption{local SSD*}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572388118-isilonAD_1it_25cpus_40bb-2.pdf}
    \caption{Isilon}\label{fig:gantt25isilon}
\end{subfigure}
    \captionsetup{belowskip=-10pt}
    \caption{Gantt charts for each storage device (App Direct Mode) processing 125 blocks of the 40$\mu$m BigBrain using 25 processes. *Some local SSD writes (approximately 5-10 writes) did not complete due to storage limitations.}\label{fig:gantt25}
\end{figure*}

As can be seen in Figure~\ref{fig:makespan-25cpus}, DRAM (App Direct mode) is the 
most efficient, with a makespan of 7.6s. Optane in Memory mode is a close second
with a makespan of approximately 10s. As caching was disabled in App Direct mode, 
Optane DCPMM in this mode is approximately 
12x slower than Optane in Memory Mode. Isilon in both Memory mode and App Direct 
mode is more than 120x slower than Optane DCPMM in Memory mode. Other than for Optane DCPMM, 
there is no significant difference between Memory mode and App Direct executions.

When comparing the makespan to the estimated read and write duration, local SSD
in both modes and Optane DCPMM in App Direct mode have longer makespans than anticipated.
Optane DCPMM was approximately 2.2x slower than expected, whereas local SSD was approximately
1.4x slower than expected, in both modes. For those where the I/O time was overestimated,
DRAM was almost 5x faster in reality compared to what was estimated. Optane DCPMM in Memory Mode
had a larger difference, with it being 3.8x faster than expected. Isilon in App Direct
mode had a smaller difference than Memory mode, with it being 1.1x faster than expected.

The total task duration breakdowns for each device (Figure~\ref{fig:stacked-25cpus})
shows that, as expected, I/O times vary greatly between devices, with DRAM having the
best total read and write speeds. Optane DCPMM in Memory Mode is very close to DRAM in speed,
whereas Optane DCPMM in App Direct mode was 7x slower in terms of reading and 23x slower in terms
of writing. Local SSD, on the other hand, exhibits similar write times as DRAM, but is
approximately 96x slower than DRAM with respect to reads. Header loading for local SSD is also 90x longer
than that of DRAM, however, it remains negligible compared to other tasks. Isilon, on the 
other hand, has non negligible header load times, with it being more than 5000x
slower than that of DRAM. Interestingly enough, the read/increment times on Isilon are
faster than that of the local SSD, although Isilon makes up for its speedier reads with 
writes that are around 184x slower than DRAM.

@@

The anticipated I/O estimates did not correctly predict the makespan for 25 parallel
processes (Figure~\ref{fig:20mksp25}). Both Optane DCPMM in Memory Mode and Isilon appear 
to be under the estimates: Optane DCPMM in Memory Mode is approximately twice as fast as the
estimates whereas Isilon is around 1.1x faster. Optane was 1.8x slower than anticipated.
Once again, Memory and App Direct Mode appear to only affect performance for Optane.

Total task duration breakdowns (Figure~\ref{fig:20total25}) show that Optane DCPMM in
App Direct Mode spends significantly more time writing than its Memory Mode counterpart.
In fact, it spends almost 12x more time writing. Isilon also appears to vary slightly
between Memory and App Direct Mode, spending around 1.4x more time reading and incrementing
in App Direct Mode.

@@

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-spark_20bb_25cpus.pdf}
    \caption{Makespan}\label{fig:20mkspspark25}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-spark-20bb_25cpus.pdf}
    \caption{Total convert/increment/write breadown}\label{fig:20totalspark:25}
\end{subfigure}
\caption{Spark incrementation application processing the 20~$\mu$m BigBrain using 25 processes. Three repetitions wereperformed}\label{fig:20spark25}
\end{figure*}

@@

A very slight performance improvement (1-1.3x faster) can be witnessed in Optane DCPMM. In 
contrast, Isilon 
experienced a slight performance decrease (1.1x slower). Whereas Isilon makespan appears to match
estimated I/O times, Optane DCPMM once again has a superior Memory mode makespan and an inferior App Direct
mode makespan (Figure~\ref{fig:20mksp96}). As previously observed with 25 processes,
the bulk of the processing time is taken up by writing for Optane DCPMM in App Direct Mode and 
Isilon in both modes (Figure~\ref{fig:20total96}). Reading takes more time in the case of Optane DCPMM 
in Memory mode.

A notable difference in the Spark execution is that Optane DCPMM in Memory and App Direct mode
exhibit the same performance (Figure~\ref{20mkspspark96}). Both of which also performed
significantly worse than the estimated I/O duration. Performance on Isilon, however, did
not really differ. The total task breakdown shows that, as expected, write times are more
significant on Isilon than on Optane DCPMM and Optane DCPMM in App Direct mode spends more time writing
than in Memory mode (Figure~\ref{fig:20totalspark96}). When comparing GNU Parallel and 
Spark implementations, Spark spent nearly half as much time writing as GNU Parallel. However,
it was found that Optane DCPMM spent more time converting the data to NiBabel than GNU Parallel spent
in loading the header.

@@

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-spark-20bb_96cpus.pdf}
    \caption{Makespan}\label{fig:20mkspspark96}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-spark-20bb_96cpus.pdf}
    \caption{Total convert/increment/write breakdown}\label{fig:20totalspark96}
\end{subfigure}
\caption{Spark incrementation application processing the 20~$\mu$m BigBrain using 96 processes. Three repetitions were performed}\label{fig:20stackedp96}
\end{figure*}
\subsection{BIDS App Example}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bids-app-example/25cores.pdf}
    \caption{25 processes}\label{fig:bm25}
\end{subfigure}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bids-app-example/96cores.pdf}
    \caption{96 processes}\label{fig:bm96}
\end{subfigure}
\caption{Makespan of BIDS App Example 25 and 96 processes on all storage devices. Three repetitions were performed}
\end{figure*}

@@

The CPU and I/O breakdowns (Figure~\ref{fig:bbd}) show that there is a significant increase
in the time spent on I/O at 96 processes when compared to 25 processes. Furthermore, there appears
to also be an increase in the amount of time spent on CPU processing. In both cases,
tmpfs spends the least amount of time on I/O and CPU, followed by local disk, Optane, and finally,
Isilon.
\section{Discussion}
\subsection{Memory vs App Direct Mode}

In general, the selection of storage mode did not affect overall performance. This can be
due to several reasons. For instance, the amount of available DRAM was abundant and exceeded 
all our dataset sizes. While total memory used by the application may have exceeded DRAM, as is the
case with the 20~$\mu$ \bigbrain, the amount of memory required by the application at any 
given moment did not exceed the amount of available DRAM. Although GNU Parallel would have
been able to load all the blocks in memory had we increased the block size, it would have been unlikely
that total available DRAM would have been exceeded. However, available DRAM could have been exceeded in
Memory Mode when using devices that leverage memory as a write cache.
In our case, only the local SSD was set up to do so.

Optane DCPMM was the only device affected by the choice of storage mode. In all cases, with the 
exception of the Spark incrementation of the 20~$\mu$ \bigbrain, Memory mode was superior to 
App Direct mode. This is a result of Optane DCPMM using DRAM as cache in Memory mode, whereas Optane
DCPMM 
was configured as a writethrough device in App Direct mode. It is believed that had Optane DCPMM been
configured to use DRAM as a writeback cache in App Direct mode, we would have observed a similar
performance between the two. Furthermore, in the 40~$\mu$ \bigbrain executions, read time on Optane DCPMM in
App Direct mode appeared to be slower than Memory mode, potentially indicating that the Optane DCPMM was able
to keep the input data cached in DRAM, whereas in App Direct Mode, the data had to be loaded from Optane DCPMM.
Therefore, it is expected the Optane DCPMM in App Direct mode may always have longer read times than Memory Mode
if the data is already cached in DRAM in Memory Mode.

The Apache Spark and GNU Parallel implementations differed vastly in makespan 
for Optane DCPMM in Memory Mode, particularly at 96 processes. After taking a closer look at
the captured Spark metrics, what significantly varied between the Spark runs on
the different storage was garbage collection, with garbage collection taking longer on Optane DCPMM 
than on Isilon. Furthermore, garbage collection was longer in Memory mode than in
App Direct mode. It is suspected that, in these cases, task duration was so short
that garbage collection could not keep up. Moreover, Optane DCPMM in Memory mode may have longer
garbage collection if old anonymous data cannot be stored in the DRAM cache due to overall memory 
requirements, therefore requiring garbage collection to occur on slower memory.

@@

For App Direct mode applications (although also applicable to Memory mode), the 
choice of using the writeback cache depends on total amount of available memory, application
memory usage and I/O. Should the application be able to use the page cache without ever being
I/O-throttled, using a write-back cache will significantly improve performance. This can 
be seen in Figure~\ref{fig:stacked-25cpus}, for instance, where the local SSD is writing
entirely to memory, significantly reducing the total write time. However, should write operations
become throttled, this may slow down the write task duration. In such cases,
it may be preferable to set them up as write-through devices. As Big Data neuroimaging applications
tend to generate temporary data files that are larger than input size, it is likely that maintaining the
devices as write-through is preferable. Furthermore, despite the fact that Optane DCPMM in App Direct Mode does
not leverage the writeback cache like the local SSD, is it still found to be superior in performance.
Therefore, while the performance of Optane DCPMM may not be enhanced by DRAM, it is still superior to that of
an SSD using the writeback cache and scalable network storage.

@@

Despite the fact that only sequential I/O was considered for the anticipated I/O times, 
Optane DCPMM in App Direct mode always took more time than its sequential estimates. The cause
of the longer-than-expected makespan seems to be due to write times. While there may be application overheads,
they should, in theory, affect Optane DCPMM in App Direct Mode as much as DRAM. While it is unknown
why Optane DCPMM writes are performing so poorly, it could be a result of filesystem configuration or 
inaccurate benchmarking of the disk. Otherwise, Optane DCPMM does not appear to scale very well with respect to
writes. Using the storage benchmarks found in Table~\ref{table:bandwidths}, it is found that with an equivalent
number of storage devices, processing the 40~$\mu$m \bigbrain using 25 processes would be faster using
local disk when compared to the real times obtained on Optane DCPMM. As there are overheads with both Optane DCPMM and local
SSD, it is uncertain that it would, in fact, be the case.

>>>>

\section{Introduction}
Neuroimaging open-data initiatives have led to extensively large repositories of
publicly available data. Such initiatives include the \bigbrain~\cite{BigBrain}, 
a one-of-a-kind $\SI{603}{\gibi\byte}$
histological image of a 65 year-old healthy human brain
at 20$\mu$m resolution; the UK Biobank~\cite{ukbiobank}, a repository expected to
contain approximately $\SI{0.2}{\peta\byte}$ of data (including various magnetic
resonance (MR) imaging modalities) from 500,000 individuals living in the UK;
the Human Connectome Project~\cite{HCP}, a repository containing MR scans from
1,200 healthy adults, expected to exceed $\SI{1}{\peta\byte}$ in size; 
and the Consortium for Reliability and Reproducibility (CoRR)~\cite{corr}, an
initiative which aggregates MR data from various centres around the world, 
of which 32 are currently available and make up about $\SI{937}{\gibi\byte}$ of
data in total.

@@

Intel Optane DC Persistent Memory Module~\cite{optanebrief} (DCPMM) is a high-performance
storage technology that resides on the DIMMs to reduce latency to the device.
The first generation of Intel Optane DCPMM offers capabities of 128~GiB,
256~GiB, and 512~GiB, enabling it to be more cost effective than DRAM storage.
There are two configuration modes, Memory mode and App Direct mode,
that enable the storage to either be accessed as an extension of volatile main memory
or as a non-volatile memory storage device\footnote{https://itpeernetwork.intel.com/intel-optane-dc-persistent-memory-operating-modes/}.

@@

The server used consisted of 12X~\SI{64}{\gibi\byte} DRAM devices,
resulting in a total of \SI{768}{\gibi\byte} of DRAM, and 12X~\SI{256}{\gibi\byte} Optane DCPMM
modules, resulting in total of \SI{3}{\tebi\byte} of Intel Optane DC persistent memory.
Other storage devices included a \SI{240}{\gibi\byte} Micron SATA SSD, of which
only \SI{149}{\gibi\byte} where available for the experiments, and a
\SI{720}{\tebi\byte} shared Dell EMC Isilon network-attached storage platform. Isilon is an nsf4 mounted
cluster (10Gb network) consisting of 5 nodes of 36 hard disk drives (HDD). The local disk
was the only storage device set up as a writeback device. Both Isilon and Optane DCPMM in
App Direct mode were configured as write-through devices, meaning they did not leverage
DRAM as a cache for writes. For Optane DCPMM, this was achieved by configuring the xfs filesystem with
direct access (DAX). DAX enables persistent memory to be byte addressable, bypassing the page cache. All storage devices were benchmarked using the following script: \url{https://github.com/ValHayot/paper-memory-storage/scripts/bench_disks.sh}.
The result of the benchmarks can be seen in Table \ref{table:bandwidths}.

@@

Memory mode leverages Optane DCPMM to extend the system's available
memory. In this mode, Optane DCPMM uses DRAM as a cache and is accessible as
volatile addressable main memory. By extending main memory, Memory mode enables the
fast access of large volumes of data. For instance, Memory mode on our server enabled use of
3~TiB of main memory, whereas it would have only had 768~GiB of main memory if Optane DCPMM could not be
leveraged. Although DRAM is used as cache by Optane DCPMM, it is not visible to the operating system.

@@

App Direct mode enables Optane DCPMM to be accessed as a high-performance storage device.
Unlike Memory mode, the OS is able to differentiate between DRAM and Optane DCPMM,
treating them as two distinct memory tiers. Optane DCPMM
does not use the DRAM as cache in App Direct mode when configured with DAX, as is our case.

Similarly to Memory mode, our experiments evaluated all available filesystems in App Direct
Mode. In other words, we evaluated DRAM (tmpfs mount), Optane DCPMM (\SI{1.5}{\tebi\byte} persistent
memory mount), local SSD and Isilon.

@@

For our experiments, we used the entire CoRR dataset, made available on \href{https://www.datalad.org/}{DataLad} and
applied Participant analysis to it. The BIDS App Example was executed using a Singularity container stored on
Isilon. As in the BigBrain Incrementation, the experiment was parallelized 
using GNU Parallel with 25 and 96 processes. The conditions were executed in both App Direct and Memory mode,
where Isilon and the local SSD were only evaluated in Memory mode. Each experiment was repeated 3x.

@@

\subsection{40~$\mu$m \bigbrain incrementation}
\subsubsection{25 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-40bb_25cpus.pdf}
    \captionsetup{belowskip=-10pt}
    \caption{Makespan of the \bigbrain Incrementation application using 25 
             processes on the 125 blocks on the 40~$\mu$m image given different
             storage configurations. Anticipated read and write duration was 
             measured by applying the perceived bandwidths Table~\ref{table:bandwidths}
             to Eq.~\ref{eq:makespan}. Three 
             repetitions were performed. *Local SSD did not complete the
         writing of the last few blocks (approximately 5-10) due to storage limitations.}\label{fig:makespan-25cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-40bb_25cpus.pdf}
    \captionsetup{belowskip=0pt}
    \caption{Total read/increment/write breakdown of \bigbrain Incrementation using 25 cpus. Three repetitions were performed. *Local SSD did not
    complete the writing of the last few blocks (approximately 5-10) due to storage limitations}\label{fig:stacked-25cpus}
\end{figure}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572380543-tmpfsAD_1it_25cpus_40bb-1.pdf}
    \caption{DRAM}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572382482-optaneAD_1it_25cpus_40bb-1.pdf}
    \caption{Optane DCPMM}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572382794-localAD_1it_25cpus_40bb-1.pdf}
    \caption{local SSD*}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572388118-isilonAD_1it_25cpus_40bb-2.pdf}
    \caption{Isilon}\label{fig:gantt25isilon}
\end{subfigure}
    \captionsetup{belowskip=-10pt}
    \caption{Gantt charts for each storage device (App Direct mode) processing 125 blocks of the 40$\mu$m BigBrain using 25 processes. *Some local SSD writes (approximately 5-10 writes) did not complete due to storage limitations.}\label{fig:gantt25}
\end{figure*}

As can be seen in Figure~\ref{fig:makespan-25cpus}, DRAM (App Direct mode) is the 
most efficient, with a makespan of 7.6s. Optane in Memory mode is a close second
with a makespan of approximately 10s. As caching was disabled in App Direct mode, 
Optane DCPMM in this mode is approximately 
12x slower than Optane in Memory mode. Isilon in both Memory mode and App Direct 
mode is more than 120x slower than Optane DCPMM in Memory mode. Other than for Optane DCPMM, 
there is no significant difference between Memory mode and App Direct executions.

When comparing the makespan to the estimated read and write duration, local SSD
in both modes and Optane DCPMM in App Direct mode have longer makespans than anticipated.
Optane DCPMM was approximately 2.2x slower than expected, whereas local SSD was approximately
1.4x slower than expected, in both modes. For those where the I/O time was overestimated,
DRAM was almost 5x faster in reality compared to what was estimated. Optane DCPMM in Memory mode
had a larger difference, with it being 3.8x faster than expected. Isilon in App Direct
mode had a smaller difference than Memory mode, with it being 1.1x faster than expected.

The total task duration breakdowns for each device (Figure~\ref{fig:stacked-25cpus})
shows that, as expected, I/O times vary greatly between devices, with DRAM having the
best total read and write speeds. Optane DCPMM in Memory mode is very close to DRAM in speed,
whereas Optane DCPMM in App Direct mode was 7x slower in terms of reading and 23x slower in terms
of writing. Local SSD, on the other hand, exhibits similar write times as DRAM, but is
approximately 96x slower than DRAM with respect to reads. Header loading for local SSD is also 90x longer
than that of DRAM, however, it remains negligible compared to other tasks. Isilon, on the 
other hand, has non negligible header load times, with it being more than 5000x
slower than that of DRAM. Interestingly enough, the read/increment times on Isilon are
faster than that of the local SSD, although Isilon makes up for its speedier reads with 
writes that are around 184x slower than DRAM.

@@

The anticipated I/O estimates did not correctly predict the makespan for 25 parallel
processes (Figure~\ref{fig:20mksp25}). Both Optane DCPMM in Memory mode and Isilon appear 
to be under the estimates: Optane DCPMM in Memory mode is approximately twice as fast as the
estimates whereas Isilon is around 1.1x faster. Optane was 1.8x slower than anticipated.
Once again, Memory and App Direct mode appear to only affect performance for Optane.

Total task duration breakdowns (Figure~\ref{fig:20total25}) show that Optane DCPMM in
App Direct mode spends significantly more time writing than its Memory mode counterpart.
In fact, it spends almost 12x more time writing. Isilon also appears to vary slightly
between Memory and App Direct mode, spending around 1.4x more time reading and incrementing
in App Direct mode.

@@

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-spark_20bb_25cpus.pdf}
    \caption{Makespan}\label{fig:20mkspspark25}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-spark_20bb_25cpus.pdf}
    \caption{Total convert/increment/write breadown}\label{fig:20totalspark:25}
\end{subfigure}
\caption{Spark incrementation application processing the 20~$\mu$m BigBrain using 25 processes. Three repetitions wereperformed}\label{fig:20spark25}
\end{figure*}

@@

A very slight performance improvement (1-1.3x faster) can be witnessed in Optane DCPMM. In 
contrast, Isilon 
experienced a slight performance decrease (1.1x slower). Whereas Isilon makespan appears to match
estimated I/O times, Optane DCPMM once again has a superior Memory mode makespan and an inferior App Direct
mode makespan (Figure~\ref{fig:20mksp96}). As previously observed with 25 processes,
the bulk of the processing time is taken up by writing for Optane DCPMM in App Direct mode and 
Isilon in both modes (Figure~\ref{fig:20total96}). Reading takes more time in the case of Optane DCPMM 
in Memory mode.

A notable difference in the Spark execution is that Optane DCPMM in Memory and App Direct mode
exhibit the same performance (Figure~\ref{fig:20mkspspark96}). Both of which also performed
significantly worse than the estimated I/O duration. Performance on Isilon, however, did
not really differ. The total task breakdown shows that, as expected, write times are more
significant on Isilon than on Optane DCPMM and Optane DCPMM in App Direct mode spends more time writing
than in Memory mode (Figure~\ref{fig:20totalspark96}). When comparing GNU Parallel and 
Spark implementations, Spark spent nearly half as much time writing as GNU Parallel. However,
it was found that Optane DCPMM spent more time converting the data to NiBabel than GNU Parallel spent
in loading the header.

@@

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-spark_20bb_96cpus.pdf}
    \caption{Makespan}\label{fig:20mkspspark96}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-spark_20bb_96cpus.pdf}
    \caption{Total convert/increment/write breakdown}\label{fig:20totalspark96}
\end{subfigure}
\caption{Spark incrementation application processing the 20~$\mu$m BigBrain using 96 processes. Three repetitions were performed}\label{fig:20stackedp96}
\end{figure*}
\subsection{BIDS App Example}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bids-app-example/25cores.pdf}
    \caption{25 processes}\label{fig:bm25}
\end{subfigure}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bids-app-example/96cores.pdf}
    \caption{96 processes}\label{fig:bm96}
\end{subfigure}
\caption{Makespan of BIDS App Example 25 and 96 processes on all storage devices. Three repetitions were performed}
\end{figure*}

@@

The CPU and I/O breakdowns (Figure~\ref{fig:bbd}) show that there is a significant increase
in the time spent on I/O at 96 processes when compared to 25 processes. Furthermore, there appears
to also be an increase in the amount of time spent on CPU processing. In both cases,
tmpfs spends the least amount of time on I/O and CPU, followed by local disk, Optane, and finally,
Isilon.
\section{Discussion}
\subsection{Memory vs App Direct mode}

In general, the selection of storage mode did not affect overall performance. This can be
due to several reasons. For instance, the amount of available DRAM was abundant and exceeded 
all our dataset sizes. While total memory used by the application may have exceeded DRAM, as is the
case with the 20~$\mu$ \bigbrain, the amount of memory required by the application at any 
given moment did not exceed the amount of available DRAM. Although GNU Parallel would have
been able to load all the blocks in memory had we increased the block size, it would have been unlikely
that total available DRAM would have been exceeded. However, available DRAM could have been exceeded in
Memory mode when using devices that leverage memory as a write cache.
In our case, only the local SSD was set up to do so.

Optane DCPMM was the only device affected by the choice of storage mode. In all cases, with the 
exception of the Spark incrementation of the 20~$\mu$ \bigbrain, Memory mode was superior to 
App Direct mode. This is a result of Optane DCPMM using DRAM as cache in Memory mode, whereas Optane
DCPMM 
was configured as a writethrough device in App Direct mode. It is believed that had Optane DCPMM been
configured to use DRAM as a writeback cache in App Direct mode, we would have observed a similar
performance between the two. Furthermore, in the 40~$\mu$ \bigbrain executions, read time on Optane DCPMM in
App Direct mode appeared to be slower than Memory mode, potentially indicating that the Optane DCPMM was able
to keep the input data cached in DRAM, whereas in App Direct mode, the data had to be loaded from Optane DCPMM.
Therefore, it is expected the Optane DCPMM in App Direct mode may always have longer read times than Memory mode
if the data is already cached in DRAM in Memory mode.

The Apache Spark and GNU Parallel implementations differed vastly in makespan 
for Optane DCPMM in Memory mode, particularly at 96 processes. After taking a closer look at
the captured Spark metrics, what significantly varied between the Spark runs on
the different storage was garbage collection, with garbage collection taking longer on Optane DCPMM 
than on Isilon. Furthermore, garbage collection was longer in Memory mode than in
App Direct mode. It is suspected that, in these cases, task duration was so short
that garbage collection could not keep up. Moreover, Optane DCPMM in Memory mode may have longer
garbage collection if old anonymous data cannot be stored in the DRAM cache due to overall memory 
requirements, therefore requiring garbage collection to occur on slower memory.

@@

For App Direct mode applications (although also applicable to Memory mode), the 
choice of using the writeback cache depends on total amount of available memory, application
memory usage and I/O. Should the application be able to use the page cache without ever being
I/O-throttled, using a write-back cache will significantly improve performance. This can 
be seen in Figure~\ref{fig:stacked-25cpus}, for instance, where the local SSD is writing
entirely to memory, significantly reducing the total write time. However, should write operations
become throttled, this may slow down the write task duration. In such cases,
it may be preferable to set them up as write-through devices. As Big Data neuroimaging applications
tend to generate temporary data files that are larger than input size, it is likely that maintaining the
devices as write-through is preferable. Furthermore, despite the fact that Optane DCPMM in App Direct mode does
not leverage the writeback cache like the local SSD, is it still found to be superior in performance.
Therefore, while the performance of Optane DCPMM may not be enhanced by DRAM, it is still superior to that of
an SSD using the writeback cache and scalable network storage.

@@

Despite the fact that only sequential I/O was considered for the anticipated I/O times, 
Optane DCPMM in App Direct mode always took more time than its sequential estimates. The cause
of the longer-than-expected makespan seems to be due to write times. While there may be application overheads,
they should, in theory, affect Optane DCPMM in App Direct mode as much as DRAM. While it is unknown
why Optane DCPMM writes are performing so poorly, it could be a result of filesystem configuration or 
inaccurate benchmarking of the disk. Otherwise, Optane DCPMM does not appear to scale very well with respect to
writes. Using the storage benchmarks found in Table~\ref{table:bandwidths}, it is found that with an equivalent
number of storage devices, processing the 40~$\mu$m \bigbrain using 25 processes would be faster using
local disk when compared to the real times obtained on Optane DCPMM. As there are overheads with both Optane DCPMM and local
SSD, it is uncertain that it would, in fact, be the case.
